---
title: 'Maternal determination of soldier proportion and paternal determination of soldier sex ratio in hybrid Reticulitermes (Isoptera: Rhinotermitidae) termite colonies'
output:
  html_document: default
  pdf_document: default
---

# packages
```{r}
install.packages("readxl")
install.packages("ggplot2")
install.packages("tidyverse")
install.packages("dplyr")
install.packages("lme4")
install.packages("car")
install.packages("multcomp")

library(readxl)
library(ggplot2)
library(tidyverse)
library(lme4)
library(car)
library(multcomp)
```

# Figure 2a: soldier sex ratio in field colonies
## * data loading
```{r}
s_sex_ratio_f <- read_excel("Dataset 1.xlsx", sheet="Figure 2a") %>% 
  as.data.frame(.)

s_sex_ratio_sum <- s_sex_ratio_f %>% 
  group_by(species) %>% 
  summarize(male = sum(number_of_male_soldiers),
            female = sum(number_of_female_soldiers)
            )
s_sex_ratio_sum

s_sex_ratio_Rs <- s_sex_ratio_sum %>% 
  filter(species == "R. speratus")
s_sex_ratio_Ra <- s_sex_ratio_sum %>% 
  filter(species == "R. amamianus")
```

## * figure
### * histogram of soldier sex ratio (R. amamianus)
```{r}
xmin <- 0
xmax <- 1
width <- 0.2
ymin <- 0
ymax <- 50
binwidth <- 0.05

par(ps=18)

Ra_hist <-
  ggplot(s_sex_ratio_f %>% filter(species == "R. amamianus"), aes(x = soldier_sex_ratio)) + 
  geom_histogram(boundary = 0, binwidth = binwidth, fill="black", color="white") +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), lim = c(0,1), expand = c(0,0)) +
  scale_y_continuous(breaks = seq(0, 10, by = 2), lim = c(0,7), expand = c(0,0)) +
  labs(x = "numerical sex ratio in soldiers (female ratio)",
       y = "number of colonies", 
       title ="") + 
  theme_classic()
Ra_hist

# save figure
ggsave(file = "Figure 2a-1.pdf", plot = Ra_hist, dpi = 300, width = 89, height = 89, units = c("mm"))
```

### * histogram of soldier sex ratio (R. speratus)
```{r}
Rs_hist <-
  ggplot(s_sex_ratio_f %>% filter(species == "R. speratus"), aes(x = soldier_sex_ratio)) + 
  geom_histogram(boundary = 0, binwidth = binwidth, fill="black", color="white") +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), lim = c(0,1), expand = c(0,0)) +
  scale_y_continuous(breaks = seq(0, 10, by = 2), lim = c(0,7), expand = c(0,0)) +
  labs(x = "numerical sex ratio in soldiers (female ratio)",
       y = "number of colonies", 
       title ="") + 
  theme_classic()
Rs_hist

# save figure
ggsave(file = "Figure 2a-2.pdf", plot = Rs_hist, dpi = 300, width = 89, height = 89, units = c("mm"))
```

## * statistical analysis
### * sex ratio bias in R. amanianus
```{r}
binom.test(c(s_sex_ratio_Ra$female,s_sex_ratio_Ra$male), p = 0.5)
```

### * sex ratio bias in R. speratus
```{r}
binom.test(c(s_sex_ratio_Rs$female,s_sex_ratio_Rs$male), p = 0.5)
```


# Figure 2b: soldier sex ratio in lab colonies (2 years old)
## * data loading
```{r}
s_sex_ratio_l <- read_excel("Dataset 1.xlsx", sheet="Fig2b, Fig3a, Fig4a") %>% 
  as.data.frame(.) %>% 
  separate(cross_type, sep = "_", c("father_species", "mother_species")) %>% 
  mutate(cross_type = paste(father_species, mother_species, sep="_"))
```

## * figure
```{r}
s_sex_ratio_l$cross_type <- factor(s_sex_ratio_l$cross_type, levels=c("Ra_Ra","Ra_Rs","Rs_Ra","Rs_Rs"))

Figure_2b <-
ggplot(s_sex_ratio_l, aes(y = soldier_sex_ratio, x = factor(cross_type))) + 
  geom_boxplot(outlier.shape = NA, 
               width=0.6 
               ) + 
  geom_jitter(group = "cross_type", 
              color = "black", 
              height = 0, 
              width = 0.1, 
              alpha = 0.5, 
              size = 1 
              ) +
  scale_y_continuous(breaks=seq(0,1,by=0.1),limits=c(0,1),expand=c(0,0)) +
  labs(y = "numerical sex ratio in soldiers (female ratio)", 
       x = "combination of parent species (male × female)",
       title ="") +
  theme_classic()
Figure_2b

# save figure
ggsave(file = "Figure 2b.pdf", plot = Figure_2b, dpi = 300, width = 89, height = 89, units = c("mm"))
```

## * statistical analysis
### * multiple comparison
```{r}
model_fig2b <- glmer(cbind(number_of_female_soldiers, number_of_male_soldiers) ~ cross_type + (1|colony_ID), family=binomial, s_sex_ratio_l)
result <- glht(model_fig2b, linfct = mcp(cross_type="Tukey"))
summary(result)
cld(result, level=0.05, decreasing = TRUE)
```

### * interaction of father and mother species
```{r}
model_fig2b_i <- glmer(cbind(number_of_female_soldiers, number_of_male_soldiers) ~ father_species * mother_species + (1|colony_ID), family=binomial, s_sex_ratio_l)
model_fig2b_i_null <- glmer(cbind(number_of_female_soldiers, number_of_male_soldiers) ~ father_species + mother_species + (1|colony_ID), family=binomial, s_sex_ratio_l)
summary(model_fig2b_i)
Anova(model_fig2b_i)
anova(model_fig2b_i, model_fig2b_i_null, test="Chi")
```


# Figure 2c: soldier sex ratio in lab colonies (2.5 years old)
## * data loading
```{r}
s_sex_ratio_l2 <- read_excel("Dataset 1.xlsx", sheet="Fig2c, Fig3b, Fig4b") %>% 
  as.data.frame(.) %>% 
  separate(cross_type, sep = "_", c("father_species", "mother_species")) %>% 
  mutate(cross_type = paste(father_species, mother_species, sep="_")) %>% 
  mutate(number_of_soldiers = number_of_male_soldiers+number_of_female_soldiers) %>% 
  mutate(soldier_sex_ratio = number_of_female_soldiers/number_of_soldiers) %>% 
  mutate(soldier_proportion = number_of_soldiers/number_of_workers)
```

## * figure
```{r}
s_sex_ratio_l2$cross_type <- factor(s_sex_ratio_l2$cross_type, levels=c("Ra_Ra","Ra_Rs","Rs_Ra","Rs_Rs"))

Figure_2c <-
ggplot(s_sex_ratio_l2, aes(y = soldier_sex_ratio, x = factor(cross_type))) + 
  geom_boxplot(outlier.shape = NA, 
               width=0.6 
               ) + 
  geom_jitter(group = "cross_type", 
              color = "black", 
              height = 0, 
              width = 0.1, 
              alpha = 0.5, 
              size = 1 
              ) +
  scale_y_continuous(breaks=seq(0,1,by=0.1),limits=c(0,1),expand=c(0,0)) +
  labs(y = "numerical sex ratio in soldiers (female ratio)", 
       x = "combination of parent species (male × female)",
       title ="") +
  theme_classic()
Figure_2c

# save figure
ggsave(file = "Figure 2c.pdf", plot = Figure_2c, dpi = 300, width = 89, height = 89, units = c("mm"))
```

## * statistical analysis
### * multiple comparison
```{r}
model_fig2c <- glmer(cbind(number_of_female_soldiers, number_of_male_soldiers) ~ cross_type + (1|colony_ID), family=binomial, s_sex_ratio_l2)
result <- glht(model_fig2c, linfct = mcp(cross_type="Tukey"))
summary(result)
cld(result, level=0.05, decreasing = TRUE)
```

### * interaction of father and mother species
```{r}
model_fig2c_i <- glmer(cbind(number_of_female_soldiers, number_of_male_soldiers) ~ father_species * mother_species + (1|colony_ID), family=binomial, s_sex_ratio_l2)
model_fig2c_i_null <- glmer(cbind(number_of_female_soldiers, number_of_male_soldiers) ~ father_species + mother_species + (1|colony_ID), family=binomial, s_sex_ratio_l2)
summary(model_fig2c_i)
Anova(model_fig2c_i)
anova(model_fig2c_i, model_fig2c_i_null, test="Chi")
```


# Figure 3a: soldier ratio in lab colonies (2 years old)
## * data loading
```{r}
s_ratio_l <- read_excel("Dataset 1.xlsx", sheet="Fig2b, Fig3a, Fig4a") %>% 
  as.data.frame(.) %>% 
  separate(cross_type, sep = "_", c("father_species", "mother_species")) %>% 
  mutate(cross_type = paste(father_species, mother_species, sep="_")) %>% 
  mutate(number_of_soldiers = number_of_male_soldiers+number_of_female_soldiers) %>% 
  mutate(soldier_sex_ratio = number_of_female_soldiers/number_of_soldiers) %>% 
  mutate(soldier_proportion = number_of_soldiers/number_of_workers)
```

## * figure
```{r}
s_ratio_l$cross_type <- factor(s_ratio_l$cross_type, levels=c("Ra_Ra","Ra_Rs","Rs_Ra","Rs_Rs"))

Figure_3a <-
ggplot(s_ratio_l, aes(y = soldier_proportion, x = factor(cross_type))) + 
  geom_boxplot(outlier.shape = NA, 
               width=0.6 
               ) + 
  geom_jitter(group = "cross_type", 
              color = "black", 
              height = 0, 
              width = 0.1, 
              alpha = 0.5, 
              size = 1 
              ) +
  scale_y_continuous(breaks=seq(0,1,by=0.025),limits=c(0,0.160),expand=c(0,0)) +
  labs(y = "numerical sex ratio in soldiers (female ratio)", 
       x = "combination of parent species (male × female)",
       title ="") +
  theme_classic()
Figure_3a

# save figure
ggsave(file = "Figure 3a.pdf", plot = Figure_3a, dpi = 300, width = 89, height = 89, units = c("mm"))
```

## * statistical analysis
### * multiple comparison
```{r}
model_fig3a <- glmer(cbind(number_of_soldiers, number_of_workers) ~ cross_type + (1|colony_ID), family=binomial, s_ratio_l)
result <- glht(model_fig3a, linfct = mcp(cross_type="Tukey"))
summary(result)
cld(result, level=0.05, decreasing = TRUE)
```

### * interaction of father and mother species
```{r}
model_fig3a_i <- glmer(cbind(number_of_soldiers, number_of_workers) ~ father_species * mother_species + (1|colony_ID), family=binomial, s_ratio_l)
model_fig3a_i_null <- glmer(cbind(number_of_soldiers, number_of_workers) ~ father_species + mother_species + (1|colony_ID), family=binomial, s_ratio_l)
summary(model_fig3a_i)
Anova(model_fig3a_i)
anova(model_fig3a_i, model_fig3a_i_null, test="Chi")
```


# Figure 3b: soldier ratio in lab colonies (2.5 years old)
## * data loading
```{r}
s_ratio_l2 <- read_excel("Dataset 1.xlsx", sheet="Fig2c, Fig3b, Fig4b") %>% 
  as.data.frame(.) %>% 
  separate(cross_type, sep = "_", c("father_species", "mother_species")) %>% 
  mutate(cross_type = paste(father_species, mother_species, sep="_")) %>% 
  mutate(number_of_soldiers = number_of_male_soldiers+number_of_female_soldiers) %>% 
  mutate(soldier_sex_ratio = number_of_female_soldiers/number_of_soldiers) %>% 
  mutate(soldier_proportion = number_of_soldiers/number_of_workers)
```

## * figure
```{r}
s_ratio_l2$cross_type <- factor(s_ratio_l2$cross_type, levels=c("Ra_Ra","Ra_Rs","Rs_Ra","Rs_Rs"))

Figure_3b <-
ggplot(s_ratio_l2, aes(y = soldier_proportion, x = factor(cross_type))) + 
  geom_boxplot(outlier.shape = NA, 
               width=0.6 
               ) + 
  geom_jitter(group = "cross_type", 
              color = "black", 
              height = 0, 
              width = 0.1, 
              alpha = 0.5, 
              size = 1 
              ) +
  scale_y_continuous(breaks=seq(0,1,by=0.025),limits=c(0,0.160),expand=c(0,0)) +
  labs(y = "numerical sex ratio in soldiers (female ratio)", 
       x = "combination of parent species (male × female)",
       title ="") +
  theme_classic()
Figure_3b

# save figure
ggsave(file = "Figure 3b.pdf", plot = Figure_3b, dpi = 300, width = 89, height = 89, units = c("mm"))
```

## * statistical analysis
### * multiple comparison
```{r}
model_fig3b <- glmer(cbind(number_of_soldiers, number_of_workers) ~ cross_type + (1|colony_ID), family=binomial, s_ratio_l2)
result <- glht(model_fig3b, linfct = mcp(cross_type="Tukey"))
summary(result)
cld(result, level=0.05, decreasing = TRUE)
```

### * interaction of father and mother species
```{r}
model_fig3b_i <- glmer(cbind(number_of_soldiers, number_of_workers) ~ father_species * mother_species + (1|colony_ID), family=binomial, s_ratio_l2)
model_fig3b_i_null <- glmer(cbind(number_of_soldiers, number_of_workers) ~ father_species + mother_species + (1|colony_ID), family=binomial, s_ratio_l2)
summary(model_fig3b_i)
Anova(model_fig3b_i)
anova(model_fig3b_i, model_fig3b_i_null, test="Chi")
```


# Figure 4a: number of workers in lab colonies (2 years old)
## * data loading
```{r}
w_number <- read_excel("Dataset 1.xlsx", sheet="Fig2b, Fig3a, Fig4a") %>% 
  as.data.frame(.) %>% 
  separate(cross_type, sep = "_", c("father_species", "mother_species")) %>% 
  mutate(cross_type = paste(father_species, mother_species, sep="_"))
```

## * figure
```{r}
w_number$cross_type <- factor(w_number$cross_type, levels=c("Ra_Ra","Ra_Rs","Rs_Ra","Rs_Rs"))

Figure_4a <-
ggplot(w_number, aes(y = number_of_workers, x = factor(cross_type))) + 
  geom_boxplot(outlier.shape = NA, 
               width=0.6 
               ) + 
  geom_jitter(group = "cross_type", 
              color = "black", 
              height = 0, 
              width = 0.1, 
              alpha = 0.5, 
              size = 1 
              ) +
  scale_y_continuous(breaks=seq(0,2000,by=100),limits=c(0,550),expand=c(0,0)) +
  labs(y = "number of workers in a colony", 
       x = "combination of parent species (male × female)",
       title ="") +
  theme_classic()
Figure_4a

# save figure
ggsave(file = "Figure 4a.pdf", plot = Figure_4a, dpi = 300, width = 89, height = 89, units = c("mm"))
```

## * statistical analysis
### * multiple comparison
```{r}
model_fig4a <- glmer(number_of_workers ~ cross_type + (1|colony_ID), family=poisson, w_number)
result <- glht(model_fig4a, linfct = mcp(cross_type="Tukey"))
summary(result)
cld(result, level=0.05, decreasing = TRUE)
```

### * interaction of father and mother species
```{r}
model_fig4a_i <- glmer(number_of_workers ~ father_species * mother_species + (1|colony_ID), family=poisson, w_number)
model_fig4a_i_null <- glmer(number_of_workers ~ father_species + mother_species + (1|colony_ID), family=poisson, w_number)
summary(model_fig4a_i)
Anova(model_fig4a_i)
anova(model_fig4a_i, model_fig4a_i_null, test="Chi")
```


# Figure 4b: number of workers in lab colonies (2.5 years old)
## * data loading
```{r}
w_number2 <- read_excel("Dataset 1.xlsx", sheet="Fig2c, Fig3b, Fig4b") %>% 
  as.data.frame(.) %>% 
  separate(cross_type, sep = "_", c("father_species", "mother_species")) %>% 
  mutate(cross_type = paste(father_species, mother_species, sep="_"))
```

## * figure
```{r}
w_number2$cross_type <- factor(w_number2$cross_type, levels=c("Ra_Ra","Ra_Rs","Rs_Ra","Rs_Rs"))

Figure_4b <-
ggplot(w_number2, aes(y = number_of_workers, x = factor(cross_type))) + 
  geom_boxplot(outlier.shape = NA, 
               width=0.6 
               ) + 
  geom_jitter(group = "cross_type", 
              color = "black", 
              height = 0, 
              width = 0.1, 
              alpha = 0.5, 
              size = 1 
              ) +
  scale_y_continuous(breaks=seq(0,2000,by=200),limits=c(0,1250),expand=c(0,0)) +
  labs(y = "number of workers in a colony", 
       x = "combination of parent species (male × female)",
       title ="") +
  theme_classic()
Figure_4b

# save figure
ggsave(file = "Figure 4b.pdf", plot = Figure_4b, dpi = 300, width = 89, height = 89, units = c("mm"))
```

## * statistical analysis
### * multiple comparison
```{r}
model_fig4b <- glmer(number_of_workers ~ cross_type + (1|colony_ID), family=poisson, w_number2)
result <- glht(model_fig4b, linfct = mcp(cross_type="Tukey"))
summary(result)
cld(result, level=0.05, decreasing = TRUE)
```

### * interaction of father and mother species
```{r}
model_fig4b_i <- glmer(number_of_workers ~ father_species * mother_species + (1|colony_ID), family=poisson, w_number2)
model_fig4b_i_null <- glmer(number_of_workers ~ father_species + mother_species + (1|colony_ID), family=poisson, w_number2)
summary(model_fig4b_i)
Anova(model_fig4b_i)
anova(model_fig4b_i, model_fig4b_i_null, test="Chi")
```


# Figure 5b: offspring survival rate
## * data loading
```{r}
survival_rate <- read_excel("Dataset 1.xlsx", sheet="Figure 5") %>% 
  as.data.frame(.) %>% 
  mutate(survival_rate = number_of_larvae_35d/20) %>% 
  separate(cross_type, sep = "_", c("father_species", "mother_species")) %>% 
  mutate(cross_type = paste(father_species, mother_species, sep="_"))
```

## * figure
```{r}
survival_rate$cross_type <- factor(survival_rate$cross_type, levels=c("Ra_Ra","Ra_Rs","Rs_Ra","Rs_Rs"))

Figure_5b <-
ggplot(survival_rate, aes(y = survival_rate, x = factor(cross_type))) + 
  geom_boxplot(outlier.shape = NA, 
               width=0.6 
               ) + 
  geom_jitter(group = "cross_type", 
              color = "black", 
              height = 0, 
              width = 0.1, 
              alpha = 0.5, 
              size = 1 
              ) +
  scale_y_continuous(breaks=seq(0,1,by=0.1),limits=c(0,1),expand=c(0,0)) +
  labs(y = "numerical sex ratio in soldiers (female ratio)", 
       x = "combination of parent species (male × female)",
       title ="") +
  theme_classic()
Figure_5b

# save figure
ggsave(file = "Figure 5b.pdf", plot = Figure_5b, dpi = 300, width = 89, height = 89, units = c("mm"))
```

## * statistical analysis
### * multiple comparison
```{r}
model_fig5a <- glmer(cbind(number_of_larvae_35d, 20 - number_of_larvae_35d) ~ cross_type + (1|colony_ID), family=binomial, survival_rate)
result <- glht(model_fig5a, linfct = mcp(cross_type="Tukey"))
summary(result)
cld(result, level=0.05, decreasing = TRUE)
```

### * interaction of father and mother species
```{r}
model_fig5a_i <- glmer(cbind(number_of_larvae_35d, 20 - number_of_larvae_35d) ~ father_species * mother_species + (1|colony_ID), family=binomial, survival_rate)
model_fig5a_i_null <- glmer(cbind(number_of_larvae_35d, 20 - number_of_larvae_35d) ~ father_species + mother_species + (1|colony_ID), family=binomial, survival_rate)
summary(model_fig5a_i)
Anova(model_fig5a_i)
anova(model_fig5a_i, model_fig5a_i_null, test="Chi")
```


# Figure 5c: offspring sex ratio
## * data loading
```{r}
w_sex_ratio <- read_excel("Dataset 1.xlsx", sheet="Figure 5") %>% 
  as.data.frame(.)

w_sex_ratio_sum <- w_sex_ratio %>% 
  group_by(cross_type) %>% 
  summarize(male = sum(number_of_male_workers),
            female = sum(number_of_female_workers)
            )
w_sex_ratio_sum

w_sex_ratio_RaRa <- w_sex_ratio_sum %>% 
  filter(cross_type == "Ra_Ra")
w_sex_ratio_RaRs <- w_sex_ratio_sum %>% 
  filter(cross_type == "Ra_Rs")
w_sex_ratio_RsRa <- w_sex_ratio_sum %>% 
  filter(cross_type == "Rs_Ra")
w_sex_ratio_RsRs <- w_sex_ratio_sum %>% 
  filter(cross_type == "Rs_Rs")
```

## * figure
```{r}
survival_rate$cross_type <- factor(survival_rate$cross_type, levels=c("Ra_Ra","Ra_Rs","Rs_Ra","Rs_Rs"))

Figure_5c <-
ggplot(survival_rate, aes(y = offspring_sex_ratio, x = factor(cross_type))) + 
  geom_boxplot(outlier.shape = NA, 
               width=0.6 
               ) + 
  geom_jitter(group = "cross_type", 
              color = "black", 
              height = 0, 
              width = 0.1, 
              alpha = 0.5, 
              size = 1 
              ) +
  scale_y_continuous(breaks=seq(0,1,by=0.1),limits=c(0,1)) +
  labs(y = "numerical sex ratio (female ratio)", 
       x = "combination of parent species (male × female)",
       title ="") +
  theme_classic()
Figure_5c

# save figure
ggsave(file = "Figure 5c.pdf", plot = Figure_5c, dpi = 300, width = 89, height = 89, units = c("mm"))
```

### * histogram of soldier sex ratio (Ra × Ra)
```{r}
xmin <- 0
xmax <- 1
width <- 0.2
ymin <- 0
ymax <- 50
binwidth <- 0.1

par(ps=18)

RaRa_hist <-
  ggplot(w_sex_ratio %>% filter(cross_type == "Ra_Ra"), aes(x = offspring_sex_ratio)) + 
  geom_histogram(boundary = 0, binwidth = binwidth, fill="black", color="white") +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), lim = c(0,1), expand = c(0,0)) +
  scale_y_continuous(breaks = seq(0, 10, by = 1), lim = c(0,3.2), expand = c(0,0)) +
  labs(x = "numerical sex ratio in workers(female ratio)",
       y = "number of colonies", 
       title ="") + 
  theme_classic()
RaRa_hist

# save figure
ggsave(file = "Figure 5d.pdf", plot = RaRa_hist, dpi = 300, width = 89, height = 89, units = c("mm"))
```

### * histogram of soldier sex ratio (Ra × Rs)
```{r}
RaRs_hist <-
  ggplot(w_sex_ratio %>% filter(cross_type == "Ra_Rs"), aes(x = offspring_sex_ratio)) + 
  geom_histogram(boundary = 0, binwidth = binwidth, fill="black", color="white") +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), lim = c(0,1), expand = c(0,0)) +
  scale_y_continuous(breaks = seq(0, 10, by = 1), lim = c(0,5.2), expand = c(0,0)) +
  labs(x = "numerical sex ratio in workers (female ratio)",
       y = "number of colonies", 
       title ="") + 
  theme_classic()
RaRs_hist

# save figure
ggsave(file = "Figure 5e.pdf", plot = RaRs_hist, dpi = 300, width = 89, height = 89, units = c("mm"))
```

### * histogram of soldier sex ratio (Rs × Ra)
```{r}
RsRa_hist <-
  ggplot(w_sex_ratio %>% filter(cross_type == "Rs_Ra"), aes(x = offspring_sex_ratio)) + 
  geom_histogram(boundary = 0, binwidth = binwidth, fill="black", color="white") +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), lim = c(0,1), expand = c(0,0)) +
  scale_y_continuous(breaks = seq(0, 10, by = 1), lim = c(0,5.2), expand = c(0,0)) +
  labs(x = "numerical sex ratio in workers (female ratio)",
       y = "number of colonies", 
       title ="") + 
  theme_classic()
RsRa_hist

# save figure
ggsave(file = "Figure 5f.pdf", plot = RsRa_hist, dpi = 300, width = 89, height = 89, units = c("mm"))
```

### * histogram of soldier sex ratio (Rs × Ra)
```{r}
RsRs_hist <-
  ggplot(w_sex_ratio %>% filter(cross_type == "Rs_Rs"), aes(x = offspring_sex_ratio)) + 
  geom_histogram(boundary = 0, binwidth = binwidth, fill="black", color="white") +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), lim = c(0,1), expand = c(0,0)) +
  scale_y_continuous(breaks = seq(0, 10, by = 1), lim = c(0,5.2), expand = c(0,0)) +
  labs(x = "numerical sex ratio in workers (female ratio)",
       y = "number of colonies", 
       title ="") + 
  theme_classic()
RsRs_hist

# save figure
ggsave(file = "Figure 5g.pdf", plot = RsRs_hist, dpi = 300, width = 89, height = 89, units = c("mm"))
```

## * statistical analysis
### * multiple comparison
```{r}
model_fig5b <- glmer(cbind(number_of_male_workers, number_of_female_workers) ~ cross_type + (1|colony_ID), family=binomial, w_sex_ratio)
result <- glht(model_fig5b, linfct = mcp(cross_type="Tukey"))
summary(result)
cld(result, level=0.05, decreasing = TRUE)
```

### * sex ratio bias in Ra × Ra cross
```{r}
binom.test(c(w_sex_ratio_RaRa$female,w_sex_ratio_RaRa$male), p = 0.5)
```

### * sex ratio bias in Ra × Rs cross
```{r}
binom.test(c(w_sex_ratio_RaRs$female,w_sex_ratio_RaRs$male), p = 0.5)
```

### * sex ratio bias in Rs × Ra cross
```{r}
binom.test(c(w_sex_ratio_RsRa$female,w_sex_ratio_RsRa$male), p = 0.5)
```

### * sex ratio bias in Rs × Rs cross
```{r}
binom.test(c(w_sex_ratio_RsRs$female,w_sex_ratio_RsRs$male), p = 0.5)
```


# Figure S1a: soldier sex ratio in 1-year-old colonies
## * data loading
```{r}
s_sex_ratio_1 <- read_excel("Dataset 2.xlsx", sheet="FigS1a") %>% 
  as.data.frame(.)

s_sex_ratio_1_sum <- s_sex_ratio_1 %>% 
  summarize(male = sum(number_of_male_soldiers),
            female = sum(number_of_female_soldiers)
            )
s_sex_ratio_1_sum
```

## * figure
```{r}
s_sex_ratio_1_fig <- s_sex_ratio_1_sum %>% 
  pivot_longer(., 
               male:female, 
               names_to = "sex", 
               values_to = "number_of_soldiers"
               )

Fig_S1a <-
ggplot(s_sex_ratio_1_fig, aes(x="", y=number_of_soldiers, fill=sex)) +
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0)
Fig_S1a

# save figure
ggsave(file = "Figure S1a.pdf", plot = Fig_S1a, dpi = 300, width = 89, height = 89, units = c("mm"))
```

## * statistical analysis
### * sex ratio bias in R. speratus
```{r}
binom.test(c(s_sex_ratio_1_sum$female,s_sex_ratio_1_sum$male), p = 0.5)
```


# Figure S1b: soldier sex ratio in 2.5- and 4.5-years-old colonies
## * data loading
```{r}
age_comp <- read_excel("Dataset 2.xlsx", sheet="FigS1b_c_d") %>% 
  as.data.frame(.) %>% 
  mutate(number_of_soldiers = number_of_male_soldiers+number_of_female_soldiers) %>% 
  mutate(soldier_sex_ratio = number_of_female_soldiers/number_of_soldiers) %>% 
  mutate(soldier_proportion = number_of_soldiers/number_of_workers)

age_2 <- age_comp %>% 
  filter(years_elapsed_since_colony_foundation == "2.5") %>% 
  summarize(male = sum(number_of_male_soldiers),
            female = sum(number_of_female_soldiers)
            )

age_4 <- age_comp %>% 
  filter(years_elapsed_since_colony_foundation == "4.5") %>% 
  summarize(male = sum(number_of_male_soldiers),
            female = sum(number_of_female_soldiers)
            )
```

## * figure
```{r}
Figure_S1b <-
ggplot(age_comp, aes(y = soldier_sex_ratio, x = factor(years_elapsed_since_colony_foundation))) + 
  geom_boxplot(outlier.shape = NA, 
               width=0.6 
               ) + 
  geom_jitter(group = "years_elapsed_since_colony_foundation", 
              color = "black", 
              height = 0, 
              width = 0.1, 
              alpha = 0.5, 
              size = 1 
              ) +
  scale_y_continuous(breaks=seq(0,1,by=0.1),limits=c(0,1),expand=c(0,0)) +
  labs(y = "numerical sex ratio in soldiers (female ratio)", 
       x = "colony age (years)",
       title ="") +
  theme_classic()
Figure_S1b

# save figure
ggsave(file = "Figure S1b.pdf", plot = Figure_S1b, dpi = 300, width = 89, height = 89, units = c("mm"))
```

## * statistical analysis
### * sex ratio bias in 2.5-years-old colonies
```{r}
binom.test(c(age_2$female,age_2$male), p = 0.5)
```

### * sex ratio bias in 4.5-years-old colonies
```{r}
binom.test(c(age_4$female,age_4$male), p = 0.5)
```

### * comparison
```{r}
model_figS1b <- glmer(cbind(number_of_female_soldiers, number_of_male_soldiers) ~ years_elapsed_since_colony_foundation + (1|colony_ID), family=binomial, age_comp)
model_figS1b_null <- glmer(cbind(number_of_female_soldiers, number_of_male_soldiers) ~ 1 + (1|colony_ID), family=binomial, age_comp)
summary(model_figS1b)
Anova(model_figS1b)
anova(model_figS1b, model_figS1b_null, test="Chi")
```


# Figure S1c: soldier proportion in 2.5- and 4.5-years-old colonies
## * figure
```{r}
Figure_S1c <-
ggplot(age_comp, aes(y = soldier_proportion, x = factor(years_elapsed_since_colony_foundation))) + 
  geom_boxplot(outlier.shape = NA, 
               width=0.6 
               ) + 
  geom_jitter(group = "years_elapsed_since_colony_foundation", 
              color = "black", 
              height = 0, 
              width = 0.1, 
              alpha = 0.5, 
              size = 1 
              ) +
  scale_y_continuous(breaks=seq(0,1,by=0.01),limits=c(0,0.062),expand=c(0,0)) +
  labs(y = "soldier proportion", 
       x = "colony age (years)",
       title ="") +
  theme_classic()
Figure_S1c

# save figure
ggsave(file = "Figure S1c.pdf", plot = Figure_S1c, dpi = 300, width = 89, height = 89, units = c("mm"))
```

## * statistical analysis
### * multiple comparison
```{r}
model_figS1c <- glmer(cbind(number_of_soldiers, number_of_workers) ~ years_elapsed_since_colony_foundation + (1|colony_ID), family=binomial, age_comp)
model_figS1c_null <- glmer(cbind(number_of_soldiers, number_of_workers) ~ 1 + (1|colony_ID), family=binomial, age_comp)
summary(model_figS1c)
Anova(model_figS1c)
anova(model_figS1c, model_figS1c_null, test="Chi")
```


# Figure S1d: number of workers in 2.5- and 4.5-years-old colonies
## * figure
```{r}
Figure_S1d <-
ggplot(age_comp, aes(y = number_of_workers, x = factor(years_elapsed_since_colony_foundation))) + 
  geom_boxplot(outlier.shape = NA, 
               width=0.6 
               ) + 
  geom_jitter(group = "years_elapsed_since_colony_foundation", 
              color = "black", 
              height = 0, 
              width = 0.1, 
              alpha = 0.5, 
              size = 1 
              ) +
  scale_y_continuous(breaks=seq(0,3000,by=500),limits=c(0,2600),expand=c(0,0)) +
  labs(y = "number of workers", 
       x = "colony age (years)",
       title ="") +
  theme_classic()
Figure_S1d

# save figure
ggsave(file = "Figure S1d.pdf", plot = Figure_S1d, dpi = 300, width = 89, height = 89, units = c("mm"))
```

## * statistical analysis
### * multiple comparison
```{r}
model_figS1d <- glmer(number_of_workers ~ years_elapsed_since_colony_foundation + (1|colony_ID), family=poisson, age_comp)
model_figS1d_null <- glmer(number_of_workers ~ 1 + (1|colony_ID), family=poisson, age_comp)
summary(model_figS1d)
Anova(model_figS1d)
anova(model_figS1d, model_figS1d_null, test="Chi")
```
